---
title: "Introduction to Tidyverse"
author: "Vamshidhar Gangu"
date: "`r Sys.Date()`"
credits: "Most of the code snippets are from R for Datascience book"
output:
  rmdformats::html_clean:
    highlight: kate
---


```{r setup, echo=FALSE, cache=FALSE}
suppressMessages(library(knitr))
suppressPackageStartupMessages(library(rmdformats))

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```
## Getting started 

The `tidyverse` is a collection of essential R packages for data science. They are quite helpful in performing most of the tedious steps of a data science project like subsetting, transforming, visualization etc

Tidyverse was created by Hadley Wickham and his team wih the aim of providing all the utilites to clean and work with data.

### "tidy" data

"Tidy data" describes a standardized approach for structuring datasets to make the downstream analyses and visualizations easy. 

the tidy data principles:

- Variables are "columns"
- Observations are "rows"
- Values go into the cells


### pipe "%>%"

part of `magrittr` package. helps to do perform many functions at the same time. the pipe makes your code read more like a sentence. 

Typical R style for select and filter would be 

```
data2 <- filter(data, department=="Computer Science")
data2_pass <-  select (data2, Marks, Country, is_pass)
```
can do nested as well:
```
data2_pass <- select(filter(data, department="Computer Science"), Marks, Country, is_pass)
```

No wrong with this but it may clutter your workspace with lots of intermediate objects. With many of these intermediate steps, it can be hard to keep track of

with pipes:

```

data2_pass <- data %>%
                filter(department=="Computer Science") %>%
                select(Marks, Country, is_pass)
```


## Tidyverse Core Packages

tidyverse can be installed direction within RStudio

```
install.packages("tidyverse")
```

The core packages includes in tidyverse:

- `readr`: fast and friendly way to read rectangular data
- `tidyr`: provides fundamental functions for data tidying
- `dplyr`: grammar for data manipulation
- `tibble`: modern data frame
- `stringr`: cohesive set of functions for string manipulation
- `forcats`: provides tools that solve common problems with factors
- `purrr`: provides tools to apply functional programming on datasets
- `ggplot2`: grammar of Graphics. 


In this part, we cover the two main core packages `tidyr` and `dplyr`

### dplyr 

gives collections of "verbs" for manipulating data frames. Each verb takes a data frame and returns a modified version of it. the idead is that complex operations can be performed by stringing together a seried of simplier operations in a pipeline

#### dyplr basics

the key dplyr functions:

- `filter` - filter rows by criteria
- `arrange` - arrange your column data (asc/desc) order
- `select` - select columns, used commonly with `starts_with`
- `mutate` - create new cols by preserving the exising variables
- `summarise` - summarize any of these functions
- `group_by` - groub different observations together and apply functions on the grouped dataset
- `join` - perform left, right, full and inner joins (similar to SQL)

#### nycflights example

use data from nycflights13 package, the flights table contains ~336K flights departed from new york city in 2013. The data comes from US Bureau of Transportation Statistics, and is documented in `?flights`

`flights`: flights data
`airlines`: airline data
`airports`: airport metadata
`planes`: plane metadata
`weather`: hourly weather data


```{r}
if(!is.element('nycflights13', installed.packages()[,1])){
  install.packages("nycflights13")
}
library("nycflights13")
library("tidyverse")
```

check the table input
```{r}
flights
```
##### select 

`select()` is used to choose a useful subset of data based on the names of the variable

```{r}
# select by name
flights %>% select(year, month, day)

# select using name range
select(flights, year:day)

# reverse select
select(flights, -(year:day))

```
other helper functions, commonly used with select

`starts_with("abc")`: to select column names starting with "abc"
`ends_with('xyz')`: matches names that ends with 'xyz' 
`contains('ijk')`: matches names that contain 'ijk'

see `?select` for more details


##### filter rows

`filter` is used to subset the data based on the condition over a column variable. Here we select all flights on Jan 1st

```{r}
jan1 <- flights %>% filter(month==1, day==1)
(dec10 <- filter(flights, month==12, day==10))
```

you can use standard suite of comparison operators, `>`, `>=`, `<`, `<=`, `!=`, `==`

```{r}
last2months <- filter(flights, month==11|month==12)
#you can use %in% for this
filter(flights, month %in% c(11,12))
```

__Excercise 1__

Q: Find flights that 
- had an arrival delay of two or more hours
- were operated by United, American or Delta

#### arrange rows

`arrange()` is used to arrange the rows based on their order. it takes a dataframe and column names to be sorted. 

```{r}
flights %>% arrange(year, month, desc(day))
```
Note: Missing Values (NA/NaN) are always sorted at the end


```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
arrange(df, desc(x))
```

__Excercise 2__

Q: select carrier and dep_delay columns from flights and sort by dep_delay

##### grouped summaries

`group_by()` changes the unit of analysis from complete dataset to individual groups. summary() is often used along with group_by to get grouped summaries

Here, we calulate the mean arraival and departure delays at different origin airports 

```{r}
flights %>%
    group_by(origin) %>%
    summarise_each(funs(mean(., na.rm=TRUE)), arr_delay, dep_delay) %>%
    head()
```
##### muatate columns

`mutate()` is used to create new variables (columns) that are functions of existing columns. 

```{r}
flights %>%
    select(year:day, ends_with("delay"), distance, air_time) %>%
    mutate(gain = dep_delay - arr_delay, speed = distance/air_time * 60)
```

use `transmutate()` if you only want to keep the new variables.


#### tidyr

it helps to tidy data into an analysis ready format. the two main functions are `gather` and `spread`

__`gather()`__ reshaping wide to long format

```{r}

wide_data <- data.frame(id = c("gene1", "gene2", "gene3", "gene4"),
        annotation = c("m protein", "k protein", "l protein", "y protein"),
        L4_control_A1 = rnorm(4, mean = 20, sd = 3), 
        C6_control_A1 = rnorm(4, mean = 20, sd = 3),
        L4_chemical_A1 = rnorm(4, mean = 20, sd = 3), 
        C6_chemical_A1 = rnorm(4, mean = 20, sd = 3) )

glimpse(wide_data)

tidy_data <- wide_data %>% gather(sample, expression, L4_control_A1: C6_chemical_A1)
tidy_data
```
__`spread()`__ opposite of gather

```{r}
wide2 <- spread(tidy_data, sample, expression)
head(wide2)
```
__`seperate()`__ and __`extract()`__
they are useful to split a character column into multiple columns. use __`unite()`__  to combine multiple columns into a single character column

```{r}
extracted_data <- tidy_data %>%
                extract(sample, into=c("genotype", "treatment", "tissue", "rep"),
                regex= "(C6|L4)_(control|chemical)_(A|B|C)(1|2|3)")
head(extracted_data)
```

### Resources

More information on tidyverse packages is availabe her [R for Data science](https://r4ds.had.co.nz/) by _Hadley Wickham_ and _Garrett Grolemund_