---
title: "Introduction to R"
author: "Vamshidhar Gangu"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: kate
---


```{r setup, echo=FALSE, cache=FALSE}
suppressMessages(library(knitr))
suppressPackageStartupMessages(library(rmdformats))

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# R Environment

### Working Directory

R typically needs a folder to store data, script and any generated plots. it also saves the history and environment by default in this **working directory**. You can get and set the working directory using `getwd()` and `setwd()`.
  


### Create a new RStudio project

You can check [here](https://bookdown.org/ndphillips/YaRrr/projects-in-rstudio.html) for Manging projects in RStudio

# R Basics

## Objects & Functions

Almost everything in R is of two things, either an object or functions. An object can be a number, a dataset, a summary statistic like a mean or statistical test. 

R object types
 
* __Scalars__ : just a single value like number/character
* __Vectors__ : collection of several scalars stored in a single object
* __Factors__ : Special type of vector used for storing categorical data and their levels
* __Lists__ : where vectors always have to contain the same type of data, a __list__ can combine just about any set of objects
* __Matrix__: contains of vectors arranged in a 2 dimensional rectangular layout. All elements have to be of the same data type (mostly numeric)
* __DataFrames__: most common format in R. it is used to store tabular data objects like spreadsheet. can contain different types of data


### Scalars

It is the simplest object type in R. It takes a single value like number/string
```{r}
# numeric scalars
a <- 10
b <- 3/100
c <- (a+b)/b
mode(c)

# character scalars 

myname <- "vamshi"
message <- 'Welcome to R course'
class(myname)

# numeric to char conv
a_char <- as.character(a)

# char to numeric conv
myname_num <- as.numeric(myname)
print(myname_num)
n = as.numeric(a_char)
print(n)

```
### Vectors

A vector is just a combination of several scalars stores in a single object. Like scalars, they can be either numeric or character but not both.

```{r}

# creating vectos using different functions

## using c() 
arr_num <- c(1,20,4,8,902)
arr_chr <- c('vamshi','nusit','nus')

# using range
arr_num_range <- 10:20
arr_chr_range <- letters[1:10]

#using other methods
arr_num_seq <- seq(10,50,2)
print(arr_num_seq)
arr_num_sample <- sample(1:1000, 20) # replace = TRUE
print(arr_num_sample)
arr_num_rep <- rep(c(1,2), times=2, each=3)
print(arr_num_rep)


# calculate the length of the vector
length(arr_num_sample)
# indexing the vector
print(arr_num_sample[3])

# calculate the mean of sample_num object using mean() function
m <- mean(arr_num_sample)
sprintf("Mean of Sample_num is :%.3f", m)

```

### Factors

A special type of vector used for saving categorical data along with its levels. Can store both strings and numerics. 
```{r}
# create a vector 
zones <- rep(c('east','west','north'), times=3, each=2)
print(zones)
print(is.factor(zones))

# converting a vector to factor 
zones_fac <- factor(zones)
print(zones_fac)
print(is.factor(zones_fac))

# create factor from scratch
fac_var <-  factor(c('male','female','female','male','male','female'), levels = c('male','female'))
print(fac_var)
```

While creating a data frame, R treats all string columns as categorical data and creates factors on it. 


### Lists

While vectors always have to contain the same type of data. a __list__ can combine just about any set of objects. It is used to store complicated objects. 
```{r}
# create few vectors

names <- c('alex','sam','vijay','kumar')
ages <- c(50,12,14,35,20)
place <- 'Singapore'

# make a list using the above three vectors
emp <- list(names, ages, place)
print(emp)
print(length(emp))

## Indexing the list
emp[[1]]
emp[[1]][2]

```

### Matrix

A __matrix__ consists of rows and columns with all the elements of same data type. Though we can create a matrix with only characters or logical values, they are not of much use. Mostly they are used with numeric elements for mathematical caluclations. it is created using `matrix()` function.
```{r}
# create a 3x3 matrix
mat_var <- matrix(1:9, byrow = TRUE, nrow = 3)
class(mat_var)

# to find the matrix dimension
dim(mat_var)
nrow(mat_var)
ncol(mat_var)

# set column names
colnames(mat_var) <- c('c1','c2','c3')
print(mat_var)

# creating a matrix from vectors using cbind (rbind can also be used)
m2 <- as.matrix(cbind(1:4, 5:8, 9:12))
print(m2)

# indexing the matrix
m2[1,2] # for slicing 1,2 element => 5
m2[1,] # for 1st row completely
m2[,2] # for selecting 2nd column
m2[-1,] # slect all rows except first

```


### Dataframes

__Data frames__ are the most common format used in R. Similar to pandas dataframe in Python. This is essentially same like a matrix but can contain different types of data. You can create a new dataframe/combine many vectors into a dataframe using `data.frame()` function. A matrix can be converted to dataframe using `as.data.frame()` function

```{r}
# create some data
name <- c('Susan','Jake','John','Ben','Jennifer')
age <- c(25,20,28,21,20)
gender <- c('F','M','M','M','F')
place <- c('NewYork','Boston','Washington DC','Las Vegas', 'Chicago')
can_play <- c(TRUE, FALSE, FALSE, TRUE, FALSE)

students <- data.frame(name, age, gender, place, can_play)
print(students)


# check the class of string column
print(students$name)
print(students$place)
print(students$gender)

# we convert the name and place to character vector from factor.
students$name <- as.character(students$name)
students$place <- as.character(students$place)

print(students$name)
print(students$place)
print(students)
```

#### Some functions on matrix and dataframe

```{r}
# view top two lines
head(students, n=2)

# view the bottom lines
tail(students)

#get the dataframe summary
summary(students)

# preview the data
str(students)

# get varaible names
names(students)

# get number of rows
nrow(students)

# get number of cols
ncol(students)
```

#### Indexing the data frame

You can use the index notation: `data_frame[row_number, column_number]` or use the `$` operator to get individual varaibles within dataframe

```{r}
# get the first row
students[1,]

# get the third column
students[,3]

# get seconf data item in third row
students[3,2]

# get the gender variable
students$gender
```

#### Some commonly used functions

__`cut`__: for grouping continous varibles 

```{r}
# splitting the students age to 2 groups
students$agegroup <- cut(students$age, 2, include.lowest=T)
print(students)
levels(students$agegroup)
```

`table`: for frequency table

```{r}
table(students$can_play)

# to express as fractions
prop.table(table(students$can_play))
```
`asDate`: Data variable
```{r}
as.Date(c("2007-06-22", "2004-02-12"))

as.Date(c("22/6/2007","13/2/2004"), format="%d/%m/%Y")
```


#### Missing Values

Missing values are denoted by NA or NaN 
- `is.na()` is used to test objects if they are NA
- `is.nan()` is used to test for NaN
- NA contains class types, like integer NA, character NA etc
- A NaN values is also NA but the converse is not true
- should be careful when combining missing values with logical expressions
```{r}
x <- c(1,2, NA, 10, 3)
is.na(x)
is.nan(x)

y <- c(1,2,NaN, NA, 4)
is.na(y)
is.nan(y)

```

removing missing values

```{r}
x[!is.na(x)]

# replace all missing values to 0
y[is.na(y)] <- 0
```

subsetting the dataframe with no missing values

```{r}
# removing missing values in data frame
students$marks <- c(99,80,NaN,100,NA)
print(students)
good <- students[complete.cases(students),]
print(good)
```


## Operators and Loops 

### Basic Operators

- Arithmetic : `+` `-` `*` `/` `^ or **`
- Logical: `<` `<=` `>` `>=` `==` `!=` `|` `&`


### Looping in R

#### `for` loops

example: count number of even numbers in a vector

```{r}

x <- 1:14
count <-0
for(i in x){
  if (i %% 2 ==0){
    count <- count +1
  }
}
print(count)
```


### `while` loops
```{r}
count <- 0
i<-1
while(i<= length(x)){
  if(x[i] %% 2==0){
    count <- count +1
  }
  i <- i+1
}
print(count)
```


### `repeat` loops
```{r}
i <- 1
count <- 0
repeat{
  if(x[i] %%2 ==0){
    count <- count + 1
  }
  i <- i + 1
  if(i > length(x)){
    break
  }
}
print(count)
```

## Readind data in R

Loading data into R might be frustation. R provides a seperate function to read almost every single type of file and can be confusing some times. 

 - tab-delimted (.txt) : `read.table()`
 - csv : `read.csv()`, `read.csv2()`
 - custom delimiter: `read.delim()`
 - excel files: `read_excel` from `readxl` package
 - json files: `fromJSON` from `rjson` package
 - xml files: `XML` package
 - log files: `read_log()`
 - FWF files: `read_fwf()`
 
 
### download data from web
 
User can download directly from web using `RCurl` pacakage 
 
```{r}

library(RCurl)
url_file <- "https://vincentarelbundock.github.io/Rdatasets/csv/datasets/iris.csv"
download.file(url_file, destfile = "iris.csv")
iris <- read.csv("iris.csv")
head(iris)
```

Most of the functions has ability to read directly from the web
```{r}
# Read a text file from the web
fromweb <- read.table(file = url_file,
                      sep = ',',
                      header = TRUE)
head(fromweb)
```

## Saving to files

Similar to read functions, there is a seperate function for writing every single type of file. MOre help can be found [here](http://www.sthda.com/english/wiki/exporting-data-from-r)

- `write.table()`, `write.csv()`, `write.delim()`

## Saving R environment

You can save the R objecta as `.RData` files. They are specific to R and can store as many objects as you'd like within a single file.

`ls()` is used to list all the variables within R environment

```{r}
ls()
```

Now we save few objects into a `.RData` object

```{r}
save(zones, students, emp, file="data_frames.RData")
```

to save all variables within a environment, use `save.image()`

```{r}
save.image(file="allenv.RData")
```

you can load the saved `.RData` using `load()` function. `rm()` is used to delete the r objects from the environment.


# To learn

learn about different [apply](https://bookdown.org/csgillespie/efficientR/programming.html#the-apply-family) functions within R

`apply`, `lapply`, `sapply`

